Measuring Perceptions and Preferences for Meritocracy

### 2021-09-03 paper.Rmd
version for SJR

---
title: "Measuring Perceptions and Preferences for Meritocracy"
css: "custom.css"
linestretch: '1.5'
link-citations: yes
author: "Juan Carlos Castillo, Julio Iturra, Luis Maldonado, Francisco Meneses & Jorge Atria <br><br>"
abstract: "Economic and social inequalities have generated growing concern and crises across contemporary societies. One of the mechanisms proposed by social sciences to explain the persistence of inequality is the belief in meritocracy, which would legitimize economic disparities based on differences in effort and talent. Despite its wide use as a concept, empirical research on meritocracy is relatively novel and characterized by diverse conceptualization and measures that make the findings and their interpretation rather confusing. Most of the studies in the area have relied upon secondary data to operationalize meritocracy, with a wide variation in the use and interpretation of the same survey items. Taking into account the extant literature that uses measures of meritocracy, this article identifies a series of drawbacks and inconsistencies within and between studies regarding conceptualization and operationalization. Based on this critical analysis, we propose a survey instrument called *Perceptions and Preferences for Meritocracy Scale*, which we test with confirmatory factor analysis using data from an online survey study (N=2,141). Our results support the proposed conceptual structure, which not only distinguishes between perceptions and preferences, but also between meritocratic and non-meritocratic dimensions. Our discussion highlights the importance of considering these different dimensions in order to advance the study of meritocracy."
output:
#  bookdown::pdf_document2:
#    template: null
#    toc: false
  bookdown::html_document2:
    number_sections: false
linkcolor: blue
bibliography:
  - input/bib/meritocracy.bib
editor_options:
  chunk_output_type: console
geometry: margin=0.78in
header-includes:
  # - \usepackage[spanish,es-tabla,es-nodecimaldot]{babel}
  - \usepackage{times}           # Times New Roman
  - \usepackage{caption}
  - \captionsetup[figure, table]{labelfont={bf},labelformat={default},labelsep=period}
  - \usepackage{graphicx}
  - \usepackage{float}
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{array}
  - \usepackage{multirow}
  - \usepackage{wrapfig}
  - \usepackage{float}
  - \usepackage{colortbl}
  - \usepackage{pdflscape}
  - \usepackage{tabu}
  - \usepackage{threeparttable}
---

@@

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = FALSE,
	fig.pos = "H",
	message = FALSE,
	warning = FALSE,
	cache = FALSE,
	out.width = "85%"
)
# knitr::opts_knit$set(base.url = "../") #relative path for .html output file
# knitr::opts_knit$set(root.dir = "../") #relative path for chunks within .rmd files
options(scipen=999)
rm(list=ls())
options(knitr.kable.NA = '')
options(knitr.graphics.error = FALSE)
Sys.setlocale("LC_ALL", "ES_ES.UTF-8")
```

@@

Economic inequality and income concentration have become an issue of growing concern in the last years. This has been expressed in a series of social movements as well as diverse critical analyses regarding the development of capitalism and its consequences [@streeckPoliticsPublicDebt2014; @piketty_Capital_2014]. In this context, the study of views, preferences and perceptions of inequality has acquired relevance in the social sciences, in topics such as redistributive preferences [@alesina_Fairness_2005; @dimick_Models_2018], the legitimization of economic inequality [@schroder_Income_2017] and the functioning of meritocracy [@duru-bellat_whos_2012; @mijs_paradox_2019; @reynolds_perceptions_2014; @atriaEconomicElitesAttitudes2020]. Within this context, the ideal of meritocracy has been strongly challenged as an unfulfilled promise of modern societies that allows the perpetuation of social inequalities [@sandel_tyranny_2020; @goldthorpe_myth_2003].

@@

Due to the role that meritocratic beliefs play in the justification of individual achievement (or failure) in contemporary societies [@hadjar_meritokratie_2008, @markovits_meritocracy_2019; @sandel_tyranny_2020], multiple studies have evaluated the relationship between meritocratic beliefs and personal and/or contextual characteristics. For example, some studies have linked meritocracy to the reinforcement of socio-economic, gender, and ethnic stereotypes [@MadeiraPrimesConsequencesSystematic2019; @girerdNeoliberalismIdeologicalBarrier2020; @premingerMeritocracyServiceEthnocracy2020], as well as the effects of meritocratic beliefs in educational [@generettStoriesWeTell2020; @owensENGINESSOCIALMOBILITY2020] and organizational contexts [@perezAdvancingCareersMerit2020; @aielloNewEvidenceFirmuniversity2019]. Most of such studies so far have used indicators from existing standard surveys (as the International Social Survey Programme - ISSP) to measure meritocracy. However, as we will show later, the concepts used as well as the instruments for  measuring meritocracy vary extensively among studies. In many cases, similar phenomena are associated with different indicators, and conversely, often different phenomena are measured with similar indicators. Such trends not only limits the comparability of studies, but also the ability to understanding the effects of meritocratic beliefs across different lines of research.

Based on a critical analysis of different approaches to the measurement of meritocracy to date, this article proposes an instrument to both measure and relate two key aspects in the study of meritocracy: perceptions and preferences. Furthermore, as a second axis of analysis, it considers the elaboration of indicators regarding meritocratic and non-meritocratic aspects, demonstrating that they are not two poles of the same continuum as some previous studies would seem to suggest. The measurement system outlined is oriented to generate an instrument as simple and brief as possible, being suitable to be used in public opinion surveys and allowing to integrate meritocratic beliefs in the study of different social phenomena.

@@

In the following we discuss four critical aspects in studies conceptualizing and measuring meritocracy and that are the basis of our scale proposal.

@@

In this measure of meritocratic beliefs, we can observe a couple of strong conceptual assumptions: effort would not depend on parental influence, and talent (as innate ability) is not meritocratic (contrary to Michael Youngs' original conceptualization). This conceptual and measurement-based assumption is found in other studies, that also assume effort to be the main and only aspect of meritocracy [@girerdNeoliberalismIdeologicalBarrier2020a; @bubakPerceptionsMeritocracyNote2019], raising the question: Is effort the only dimension behind the concept of meritocracy? The question of whether talent is or is not considered meritocratic is certainly an interesting topic to discuss from a philosophical point of view, but for those working on empirical studies it should be something that face empirical scrutiny. Besides, the consideration of talent as part of meritocracy certainly opens some interesting avenues of research, for instance in studies showing that for the elites meritocracy is more related to talent, whereas effort is something more characteristic of meritocracy of the middle and low classes ([citar Atria ABS]).

@@

Several approaches to the empirical study of meritocracy based on public opinion surveys, refer to the concept of _beliefs_, however different meanings and operationalizations usually underlie this concept. To illustrate this point, lets consider the most commonly used items for operationalizing meritocratic beliefs in survey research (as ISSP), which is the "reasons to get ahead" battery. It consists in a series of items asking "how important you think it is for getting ahead in life" and then lists several factors, such as effort, education, parental wealth, and contacts. Another version of this same battery used in several surveys - sometimes along the previous one - asks about "how important you think it _should_ be ...", then listing the same concepts.  Therefore, the question raised here is: Which one of both is a "belief"?  What _is_ or what it _should_ be?

The term belief has an ambiguous character in the literature, conceived as "idea-elements" by @converseNatureBeliefSystems1964 or "considerations" by @zallerNatureOriginsMass1992. As @kluegelBeliefsInequalityAmericans2017 pointed out about the scope of beliefs: "This usage encompasses such more specific social-psychological concepts as values, perceptions, and attitudes" (p.30). Therefore, beliefs cover almost anything related to subjective factors. To this regard, a relevant distinction in the field of inequality beliefs was made by @janmaatSubjectiveInequalityReview2013: "Perceptions refer to subjective estimates of existing inequality (i.e. thoughts about what is). Beliefs are here defined as normative ideas about just inequality (i.e. thoughts about what should be)"(p.359). Several papers dealing with meritocracy use the term beliefs (i.e. what should be), while actually referring to perceptions (i.e. what is). This occurs for instance in @reynolds_perceptions_2014, in which the term beliefs is used to talk about perceptions, whereas other authors use general terms such as attitudes [@kunovich_systems_2007]. The first attempt to shed light on this issue was made by @duru-bellat_whos_2012, who used the question "how important should the number of years spent in education and training be in deciding how much money people ought to earn?" as a proxy for "desired" meritocracy (beliefs). They then determined "perceived" meritocracy, using the questions: “Would you say that in your country, people are rewarded for their efforts?” and “... people are rewarded for their skills?”.

Is the belief in meritocracy a perception or a preference with normative meaning? In order to expand the analytical conceptual framework, we believe that both dimensions should be included in analyses, as proposed by @duru-bellat_whos_2012. This opens up the possibility of analyzing whether perceptions and preferences are actually related (i.e. have a high correlation), or whether they are independent aspects of the same phenomenon. As @son_hing_merit_2011-1 have pointed out, "People can believe that outcomes ought to be distributed on the basis of merit and yet vary in their perceptions of whether this is how society currently operates" (p. 435). In other words, normative beliefs should be considered while taking perceptions into account: a strong normative belief in meritocracy may mean something totally different to someone perceiving high meritocracy, when compared to someone perceiving low meritocracy. To avoid the confusion generated by the term "belief", we propose the terms meritocratic preferences ("what should be"), and meritocratic perceptions ("what is"), as they better reflect the two facets of meritocracy under scrutiny [@castillo_meritocracia_2019].

@@

A number of studies have dealt with aspects opposed to meritocracy, as for instance the use of personal advantages (as contacts or having a wealthy family) to get ahead in life. For instance, @kunovich_systems_2007 used an items’ battery listing a number of factors in relation to “How important each should be in deciding pay...” (as @duru-bellat_whos_2012 for desired meritocracy). They consider factors such as education and responsibility as meritocratic, giving them a value of 1 if considered “essential” in the scale response, whereas factors such as having a family or children are valued 1 when rated as "not important at all" (i.e. reverse coded). The assumption behind this approach is that rejecting a supposed non-meritocratic aspect (as having family and/or children) implies a stronger belief in meritocracy. A similar approach of reverse-coding non-meritocratic items was taken by @newman_false_2015, using the same principles applied in the "Preference for the Merit Principle Scale" [@davey_preference_1999].

The assumption that meritocratic and non-meritocratic elements are poles of the same continuum was tested by @reynolds_perceptions_2014 using the "get ahead" perceptions' battery items mentioned above. They considered education, ambition, and hard work as meritocratic, and other factors such as family wealth and connections as non-meritocratic. Despite making and proving this distinction, however, the authors end up subtracting one dimension from the other, thus coming back to the assumption that they are two poles of the same continuum as @kunovich_systems_2007 did. Similarly, @roex_attitudes_2019 used ISSP indicators for perceived meritocracy and non-meritocracy to build a single score by reverse coding the non-meritocratic items. Therefore, the treatment of non-meritocratic items has been rather inconsistent across studies and the assumption that they are the simple opposite of meritocracy requires further empirical assessment.

@@

Finally, most studies in meritocracy so far have not accounted for the issue of measurement error [@brownBookReviewTimothy2010; @bollenStructuralEquationsLatent1989] (Ansolabehere, S., Rodden, J., & Snyder Jr, J. M., 2008), as they use single indicators and/or simple average indexes for measuring meritocracy. Such strategy assumes that the latent construct is measured perfectly (i.e. no error or residual variance) by the indicators chosen, going as far as to propose that "... In choosing this strategy of index construction, we argue that support for meritocracy is not a latent variable" [@kunovich_systems_2007 p.653-654]. Some advances were made by @reynolds_perceptions_2014 through conducting a principal component analysis of meritocratic and non-meritocratic dimensions, but somewhat contradictorily, they chose a sum index despite proving a multidimensional latent structure.

## An instrument proposal {-#instrumentprop}

Based on the previous assumptions and limitations of the measurements of meritocracy, we propose and test an instrument with the following characteristics:

@@

```{r merit-model, echo=FALSE, fig.cap = "Model of perception and preferences for meritocracy and non-meritocracy", fig.align='center'}
knitr::include_graphics("input/images/generalf.png")
```
The columns "Perceptions" and "Preferences" represent the distinction between these two concepts, usually confused under the label "beliefs" [@castillo_meritocracia_2019]. Perceptions refer to the extent to which people observe that meritocracy functions or applies in their society, which relates to items such as "I think hard work is important to get ahead in society". Preferences refer to normative expectations that are usually linked to a "should" expression (e.g. whether hard work should be related to payment). The rows in Figure \@ref(fig:merit-model) consider the distinction between meritocratic and non-meritocratic dimensions [@reynolds_perceptions_2014]. These aspects have usually been treated as different ends of the same continuum in previous research, an assumption that requires empirical scrutiny. Non-meritocratic elements usually refer to the use of personal contacts or family advantages to get ahead in life.

Regarding the selection of indicators, most are taken or adapted from previous studies for the sake of comparability. For meritocratic indicators, we use effort and talent as the main components of the traditional concept of merit as defined by @young_rise_1962, whereas for non-meritocratic dimensions, we use wealthy parents and good contacts. Descriptions of the specific items are outlined in the methodology section.

The research hypotheses behind this conceptualization and measurement model are as follows:

* $H_1$. The perception of meritocracy is a latent variable based on indicators of the importance attributed to talent and effort to get ahead in life.

* $H_2$. The perception of non-meritocracy is a latent variable that derives from  indicators related to agreement with the statement that people with contacts and rich parents manage to get ahead.

* $H_3$. Meritocratic preferences are a latent variable based on the normative value of effort and talent.

* $H_4$. Non-meritocratic preferences are a latent variable based on the normative value of the use of personal contacts and having wealthy parents.

@@

table_format2 = if(is_html_output()) {
  T
} else if(is_latex_output()) {
  F
}
```

The analysis is organized in three studies. Study 1 serves as a background to our instrument proposal by attempting to operationalize the meritocracy dimensions with available secondary data. For this we use data from the International Social Survey Programme (ISSP). Although we are aware that the available indicators in this dataset do not allow a precise operationalization of the conceptual model, we know that this dataset is the most widely used in comparative meritocracy studies (as well as overall subjective inequality) and therefore it is certainly useful to look at its potentials and limitations to this regard. Study 2 is the core of this paper and consists of the proposal of a new scale for measuring meritocratic perceptions and preferences. Finally, in Study 3 we perform additional validity analysis of the meritocracy scale.

@@

The data is the last available wave of the social inequality module from the International Social Survey Programme (ISSP), which is the most specialized international comparative survey in perceptions, attitudes and beliefs about inequality related issues. This wave corresponds to the year 2009 and covers attitudes towards a series of topics dealing with social inequality across 41 countries and 56021 individuals. Although there is data available from this module for previous waves (1987, 1992 and 1999), unfortunately there are several variables that are important for this research that were not included, particularly in 1987 and 1999, reason why we only use the 2009 wave.

@@

There are a series of indicators in the ISSP survey that in the following we attempt to classify in our meritocracy measurement scheme. However, it is important to mention  that the items were not originally designed as a specific measure of meritocracy. Still, as this database is one of the most widely used for international social research, many of the items have been used in previous research on meritocracy (and certainly will keep being used in the future). Therefore, before proposing a new measure of meritocracy we  give a first test to our proposed conceptualization using available data that not only covers several countries but also is validated by the research community.

The variables are presented below organized in perceptions and preferences:

- _Perception of meritocracy/non-meritocracy_: for operationalizing perceptions the closest set of ISSP's indicators comes from the question asking about perceptions for opportunities to get ahead, which are usually considered as "meritocratic beliefs" in previous research. The general heading of the battery is: _"To begin we have some questions about opportunities for getting ahead. Please tick one box for each of these to show how important you think it is for getting ahead in life."_ This is followed by a list of statements to be answered in a scale with the following levels numbered from 1 to 5: essential, very important, fairly important, not very important, not important at all.

The classification is based on criteria of internal motivation (meritocratic) and structural constraints (non-meritocratic). There were two items from the battery that were excluded from the analysis as they would not fit into the classification. The first one was "having good education yourself", since it was not clear whether this is could be due to individual motivation or system opportunities, and the second was "giving bribes", as introduced elements of criminality that were beyond a non-meritocratic perception.

- Preferences for meritocracy-non meritocracy: for the operationalization of normative preferences we used a list of items related to reasons for pay. The ISSP question was: _In deciding how much people ought to earn, how important should each of these things be, in your opinion_, rated in the same _essential-non important at all_ scale as the questions for meritocratic preferences.

Table \@ref(tab:table-issp) below summarise the items classified according to the proposal based on perception and preference regarding to the meritocratic/non-meritocratic dimensions:

@@

@@

The estimation was performed using Confirmatory Factor Analysis (CFA). CFA was conducted using the `lavaan` R package (version 0.6-3; Rosseel, 2020), with diagonally weighted least squares (DWLS) estimation due to the items’ ordinal level of measurement (Kline, 2016; Rosseel, 2020). As recommended by Brown (2008), we assessed model fit by jointly considering the comparative fit index and Tucker-Lewis Index (CFI and TLI; acceptable fit > 0.95), Root of the average squared residual approximation (RMSEA; acceptable fit < 0.08),  Chi-square: (p-value; acceptable fit > 0.05, and Chi-square ratio > 3).

@@

<!-- (here summary descriptive table - @julio) -->

<!-- Table \@ref(tab:desc-issp) -->

@@

Figure \@ref(fig:likert-issp) shows the distribution of responses across the indicators for each perception and preference in the meritocratic and non-meritocratic dimensions. On one side, we see that there is a certain degree of importance attributed to factors such as hard work and ambition in the process of getting ahead, concentrating 94.9% and 92.8% in the fairly important to essential categories.

By contrast, for the non-meritocratic aspects, parental background showed a more dispersed opinion, where 61.2% considers that coming from a wealthy family is important, and 76% considers that having educated parents is important.  On the other hand, the background dimension shows that people' opinions are generally in the opposite sense, in which 63.7% consider that a person's racial origin is not very important or not important at all in the process of getting ahead in life, while 35% consider it important, and this is nearly identical regarding the importance of gender. Lastly, we can observe differences in the networks dimension, in which 85.7% consider it to be important to know the right people in the process of getting ahead in life, contrasting with 50% who consider political connections to be important.

Regarding normative preferences, we observe how well a person performs his or her job and amount of effort involved concentrates 98% and 96.9% of the responses in the categories that attribute importance to these characteristics when evaluating the pay that people deserve. Finally, in relation to structural constrains, 83% attribute importance to the fact that supporting a family should be a relevant factor with respect to the pay obtained, while having children to support translates into 76.4% of importance in this dimension.

@@

This section presents the results of the confirmatory factor analysis. First, a four-dimensional model for meritocratic and non-meritocratic perception and preferences is estimated, considering all indicators of the non-meritocratic perception dimension as a single latent variable. Then, we estimate a model that follows the proposed four latent variable measurement approach, however, the non-meritocratic perception dimension is assumed as a second-order factor for the first order latent variables parental attributes, racial and gender background and networks.

@@

```{r include=FALSE}

# if (knitr::is_html_output()) {tb <- table_nums("tb",cap)} else {tb <- cap}
```

>>>>

---
title: "Measuring Perceptions and Preferences for Meritocracy"
css: "custom.css"
linestretch: '1.5'
link-citations: yes
author: "Juan Carlos Castillo, Julio Iturra, Luis Maldonado, Francisco Meneses & Jorge Atria <br><br>"
abstract: "Economic and social inequalities have generated growing concerns and crises across contemporary societies. One of the mechanisms proposed by social sciences to explain the persistence of inequality is the belief in meritocracy, which would legitimize economic disparities based on differences in effort and talent. Despite its wide use as a concept, empirical research on meritocracy is relatively novel and characterized by diverse conceptualization and measures that make the findings and their interpretation rather inconsistent. Most of the studies in the area have relied upon secondary data to operationalize meritocracy, with a wide variation in the use and interpretation of the same survey items. Taking into account the extant literature that uses measures of meritocracy, this article identifies a series of drawbacks and inconsistencies within and between studies regarding conceptualization and operationalization. Based on this critical analysis, we propose a conceptual framework for studying perceptions and preferences for meritocracy and non-meritocracy, which is then tested through confirmatory analysis using ISSP (International Social Survey Programme) data as well as a novel survey designed specifically with this purpose(N=2,141). Our results support the conceptual framework and its operationalizarion, although with a better fit for the proposed scale than for the ISSP survey. Our discussion highlights the importance of considering different dimensions in order to advance in the study of meritocracy."
output:
#  bookdown::pdf_document2:
#    template: null
#    toc: false
  bookdown::html_document2:
    number_sections: false
linkcolor: blue
bibliography:
  - input/bib/meritocracy.bib
editor_options:
  chunk_output_type: console
geometry: margin=0.78in
header-includes:
  # - \usepackage[spanish,es-tabla,es-nodecimaldot]{babel}
  - \usepackage{times}           # Times New Roman
  - \usepackage{caption}
  - \captionsetup[figure, table]{labelfont={bf},labelformat={default},labelsep=period}
  - \usepackage{graphicx}
  - \usepackage{float}
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{array}
  - \usepackage{multirow}
  - \usepackage{wrapfig}
  - \usepackage{float}
  - \usepackage{colortbl}
  - \usepackage{pdflscape}
  - \usepackage{tabu}
  - \usepackage{threeparttable}
---

@@

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning = FALSE,message = FALSE, cache = FALSE,out.width = '85%',fig.pos= "H")
# knitr::opts_knit$set(base.url = "../") #relative path for .html output file
# knitr::opts_knit$set(root.dir = "../") #relative path for chunks within .rmd files
options(scipen=999)
rm(list=ls())
options(knitr.kable.NA = '')
options(knitr.graphics.error = FALSE)
Sys.setlocale("LC_ALL", "ES_ES.UTF-8")
```

@@

Economic inequality and income concentration have become topics of growing concern over the last years, acquiring even more salience in times of pandemic and economic crises. This has led to a series of social movements in different societies as well as diverse critical analyses regarding the development of capitalism and its consequences [@streeckPoliticsPublicDebt2014; @piketty_Capital_2014]. In this context, the study of views, preferences and perceptions of inequality has acquired relevance in the social sciences, in topics such as redistributive preferences [@alesina_Fairness_2005; @dimick_Models_2018], the legitimization of economic inequality [@schroder_Income_2017] and the functioning of meritocracy [@duru-bellat_whos_2012; @mijs_paradox_2019; @reynolds_perceptions_2014; @atriaEconomicElitesAttitudes2020]. Within this context, the ideal of meritocracy has been strongly challenged as an unfulfilled promise of modern societies that allows the perpetuation of social inequalities [@sandel_tyranny_2020; @goldthorpe_myth_2003].

@@

Due to the role that meritocratic beliefs play in the justification of individual achievement (or failure) in contemporary societies [@hadjar_meritokratie_2008, @markovits_meritocracy_2019; @sandel_tyranny_2020], multiple studies have evaluated the relationship between meritocratic beliefs and personal and/or contextual characteristics. For example, some studies have linked meritocracy to the reinforcement of socio-economic, gender, and ethnic stereotypes [@MadeiraPrimesConsequencesSystematic2019; @girerdNeoliberalismIdeologicalBarrier2020; @premingerMeritocracyServiceEthnocracy2020], as well as the effects of meritocratic beliefs in educational [@generettStoriesWeTell2020; @owensENGINESSOCIALMOBILITY2020] and organizational contexts [@perezAdvancingCareersMerit2020; @aielloNewEvidenceFirmuniversity2019]. Most of such studies so far have used indicators from existing standard social surveys, such as the International Social Survey Programme (ISSP), to measure meritocracy. However, as we will show later, the concepts used as well as the instruments for  measuring meritocracy vary extensively among studies. In many cases, similar phenomena are associated with different indicators, and conversely, often different phenomena are measured with similar indicators. Such trends not only limits the comparability of studies, but also the ability to understanding the effects of meritocratic beliefs across different disciplines and lines of research.

Based on a critical analysis of different approaches to the measurement of meritocracy to date, this article proposes an instrument to both measure and relate two main aspects in the study of meritocracy: perceptions and preferences. Furthermore, it distinguishes between meritocratic and non-meritocratic dimensions, as they are not necessarily two poles of the same continuum as previous studies seem to suggest. This conceptual framework is oriented to develop a measurement instrument as simple and brief as possible, being suitable to be used in public opinion surveys and allowing to integrate meritocratic beliefs in the study of different social phenomena.

@@

In the following we discuss four critical aspects in studies conceptualizing and measuring meritocracy and that are the basis of the conceptual and measurement proposal.

@@

In this conceptual and measurement approach of meritocratic beliefs, we can observe a couple of strong conceptual assumptions: a) effort would not depend on parental influence, and b) talent (as innate ability) is not meritocratic (contrary to Michael Youngs' original conceptualization). This conceptual and measurement-based assumption is found in other studies, that also assume effort to be the main and only aspect of meritocracy [@girerdNeoliberalismIdeologicalBarrier2020a; @bubakPerceptionsMeritocracyNote2019], raising the question: Is effort the only dimension behind the concept of meritocracy? The question of whether talent (as inteligence and ability) is or is not considered meritocratic is certainly an interesting topic to discuss from a philosophical point of view, but for those working on empirical studies it should face empirical scrutiny. Besides, the consideration of talent as part of meritocracy certainly opens some interesting avenues of research, for instance in studies showing that for the elites meritocracy is more related to talent, whereas effort is something more characteristic of meritocracy of the middle and low classes ([### citar Atria ABS]).

@@

Several approaches to the empirical study of meritocracy based on public opinion surveys refer to the concept of _beliefs_, but with wide differences in meanings and operationalizations. To illustrate this point, lets consider the most commonly used items for operationalizing meritocratic beliefs in survey research (as ISSP), which is the "reasons to get ahead" battery. It consists in a series of items asking "how important you think it is for getting ahead in life" and then lists several factors, such as effort, education, parental wealth, and contacts. Another version of this same battery used in other surveys - sometimes together with the previous one - asking about "how important you think it _should_ be ...", and then listing the same concepts.  Therefore, the question raised here is: Which one of both is a "belief": what _is_ or what it _should_ be?

The term belief has an ambiguous character in the literature, conceived as "idea-elements" by @converseNatureBeliefSystems1964 or "considerations" by @zallerNatureOriginsMass1992. As @kluegelBeliefsInequalityAmericans2017 pointed out about the scope of beliefs: "This usage encompasses such more specific social-psychological concepts as values, perceptions, and attitudes" (p.30). Therefore, beliefs cover almost anything related to subjective factors. To this regard, a relevant distinction in the field of inequality beliefs was made by @janmaatSubjectiveInequalityReview2013: "Perceptions refer to subjective estimates of existing inequality (i.e. thoughts about what is). Beliefs are here defined as normative ideas about just inequality (i.e. thoughts about what should be)"(p.359). Several papers dealing with meritocracy use the term beliefs (i.e. what should be), while actually referring to perceptions (i.e. what is). This occurs for instance in @reynolds_perceptions_2014, in which the term beliefs is used to talk about what @janmaatSubjectiveInequalityReview2013 refer as perceptions, whereas other authors use general terms such as attitudes [@kunovich_systems_2007]. The first attempt to shed light on this issue in meritocracy research was made by @duru-bellat_whos_2012, who used the question "how important should the number of years spent in education and training be in deciding how much money people ought to earn?" as a proxy for "desired" meritocracy (normative beliefs). They then determined "perceived" meritocracy, using the questions: “Would you say that in your country, people are rewarded for their efforts?” and “... people are rewarded for their skills?”.

Is the belief in meritocracy a perception or a preference with normative meaning? In order to expand the analytical conceptual framework, we believe that both dimensions should be included as proposed by @duru-bellat_whos_2012. This opens up the possibility of analyzing whether perceptions and preferences are actually related (i.e. have a high correlation), or whether they are independent aspects of the same phenomenon. As @son_hing_merit_2011-1 have pointed out, "People can believe that outcomes ought to be distributed on the basis of merit and yet vary in their perceptions of whether this is how society currently operates" (p. 435). In other words, normative beliefs should be considered while taking perceptions into account: a strong normative belief in meritocracy may mean something totally different to someone perceiving high meritocracy, when compared to someone perceiving low meritocracy. To avoid the confusion generated by the term "belief", we propose the terms meritocratic preferences ("what should be"), and meritocratic perceptions ("what is"), as they better reflect the two facets of meritocracy under scrutiny [@castillo_meritocracia_2019].

@@

Some research in meritocracy considers aspects usually opposed to effort and talent for personal achievement, as for instance the use of personal advantages (as contacts or having a wealthy family) to get ahead in life. For instance, @kunovich_systems_2007 used an items’ battery listing a number of factors in relation to “How important each should be in deciding pay...” (as @duru-bellat_whos_2012 for desired meritocracy). They consider factors such as education and responsibility as meritocratic, giving them a value of 1 if considered “essential” in the scale response, whereas factors such as having a family or children are valued 1 when rated as "not important at all" (i.e. reverse coded). The assumption behind this approach is that rejecting a supposed non-meritocratic aspect (as having family and/or children) implies a stronger belief in meritocracy. A similar approach of reverse-coding non-meritocratic items was taken by @newman_false_2015, using the same principles applied in the "Preference for the Merit Principle Scale" [@davey_preference_1999].

The assumption that meritocratic and non-meritocratic elements are poles of the same continuum was tested by @reynolds_perceptions_2014 using the "get ahead" perceptions' battery items mentioned above. They considered education, ambition, and hard work as meritocratic, and other factors such as family wealth and connections as non-meritocratic. Despite making and proving this distinction, however, the authors end up subtracting one dimension from the other, thus coming back to the assumption that they are two poles of the same continuum as @kunovich_systems_2007 did. Similarly, @roex_attitudes_2019 used ISSP indicators for perceived meritocracy and non-meritocracy to build a single score by reverse coding the non-meritocratic items. Therefore, the treatment of non-meritocratic items has been rather inconsistent across studies and the assumption that they are the simple opposite of meritocracy certainly requires further empirical assessment.

@@

Finally, most studies in meritocracy so far have not properly considered the issue of latent structures and measurement error [@brownBookReviewTimothy2010; @bollenStructuralEquationsLatent1989] (Ansolabehere, S., Rodden, J., & Snyder Jr, J. M., 2008), as they mostly use single indicators and/or simple average indexes for measuring meritocracy. Such strategy assumes that the latent construct is measured perfectly (i.e. no error or residual variance) by the selected indicators, going as far as to propose that "... In choosing this strategy of index construction, we argue that support for meritocracy is not a latent variable" [@kunovich_systems_2007 p.653-654]. Although some advances were made by @reynolds_perceptions_2014 through conducting a principal component analysis of meritocratic and non-meritocratic dimensions, somewhat surprisingly they finally chose to build a sum index despite proving a multidimensional latent structure.

## Conceptual proposal for studying meritocracy {-#instrumentprop}

Based on the previous assumptions and limitations identified in the empirical study of  meritocracy, we propose a conceptual and measurement framework with the following characteristics:

@@

```{r merit-model, echo=FALSE, fig.cap = "Model of perception and preferences for meritocracy and non-meritocracy", fig.align='center'}
knitr::include_graphics("input/images/generalf.png")
```
The columns "Perceptions" and "Preferences" represent the distinction between these two concepts, usually confused under the label "beliefs" [@castillo_meritocracia_2019]. Perceptions refer to the extent to which people observe that meritocracy functions or applies in their society, which relates to items such as "Hard work is important to get ahead in society". Preferences refer to normative expectations that are usually linked to a "should" expression (e.g. whether hard work should be related to payment). The rows in Figure \@ref(fig:merit-model) consider the distinction between meritocratic and non-meritocratic dimensions [@reynolds_perceptions_2014], usually treated as different ends of the same continuum in previous research. Non-meritocratic elements refer to the use of resources such as personal contacts or family advantages to get ahead in life.

@@

  table_format2 = if(is_html_output()) {
    T
  } else if(is_latex_output()) {
    F
  }

The analysis is organized in three studies. Study 1 serves as a background to our instrument proposal by attempting to operationalize the meritocracy dimensions with available secondary data from the inequality module of the International Social Survey Programme (ISSP). Although we are aware that the available indicators in this dataset do not allow a precise operationalization of the conceptual model, we know that it is the most widely used data in comparative meritocracy studies (as well as overall subjective inequality) and therefore it is certainly useful to look at its potentials and limitations to this regard. Study 2 is the core of this paper and consists of the proposal of a new scale for measuring meritocratic perceptions and preferences. Finally, in Study 3 we perform additional validity analysis of the meritocracy scale.

@@

The data cooresponds to the last available wave of the social inequality module from the International Social Survey Programme (ISSP), which is the most specialized international comparative survey in perceptions, attitudes and beliefs about inequality related issues. This wave corresponds to the year 2009 and covers attitudes towards a series of topics dealing with social inequality across 41 countries and 56021 individuals. Although there is data available from this module for previous waves (1987, 1992 and 1999), unfortunately there are several variables that are important for this research that were not included, particularly in 1987 and 1999, reason why we only use the 2009 wave (during the development of this research the 2019 wave became available, but so far only with a restricted set of countries).

@@

There are a series of indicators in the ISSP survey that in the following we attempt to classify in our meritocracy measurement scheme. However, it is important to mention  that the items were not originally designed as a specific measure of meritocracy. The variables are presented below organized in perceptions and preferences:

@@

- _Perception of meritocracy/non-meritocracy_: for operationalizing perceptions the closest set of ISSP's indicators comes from the question asking about perceptions for opportunities to get ahead, which are usually considered as "meritocratic beliefs" in previous research. The general heading of the battery is: _"To begin we have some questions about opportunities for getting ahead. Please tick one box for each of these to show how important you think it is for getting ahead in life."_ This is followed by a list of statements to be rated from 1 to 5: essential, very important, fairly important, not very important, not important at all.

The classification of the items is based on criteria of internal motivation (meritocratic) and structural constraints (non-meritocratic). There were two items from the battery that were excluded from the analysis as they would not fit into the classification. The first one was "having good education yourself", since it was not clear whether this is could be due to individual motivation or system opportunities, and the second was "giving bribes", as introduced elements of criminality that were beyond a non-meritocratic perception.

- Preferences for meritocracy-non meritocracy: for the operationalization of normative preferences we used a list of items related to reasons for pay battery. The ISSP question was: _In deciding how much people ought to earn, how important should each of these things be, in your opinion_, rated in the same _essential-non important at all_ scale (1 to 5) as the questions for meritocratic perceptions.

@@

The estimation was performed using Confirmatory Factor Analysis (CFA). CFA was conducted using the `lavaan` R package (version 0.6-3; ### Rosseel, 2020), with diagonally weighted least squares (DWLS) estimation due to the items’ ordinal level of measurement (### Kline, 2016; Rosseel, 2020). As recommended by ### Brown (2008), we assessed model fit by jointly considering the comparative fit index and Tucker-Lewis Index (CFI and TLI; acceptable fit > 0.95), Root of the average squared residual approximation (RMSEA; acceptable fit < 0.08),  Chi-square: (p-value; acceptable fit > 0.05, and Chi-square ratio > 3).

@@

@@

Figure \@ref(fig:likert-issp) shows the distribution of responses across the selected items in their corresponding dimensions. On the one side, we see that there is a high degree of importance attributed to factors such as hard work and ambition in the process of getting ahead, concentrating 94.9% and 92.8% in the fairly important to essential categories. By contrast, the perception non-meritocratic aspects is rated lower that the meritocratic ones, particularly for gender, race and political connections. Regarding normative preferences, we observe that the meritocratic ones are deemed as important (from fairly to essential) for nearly the whole sample, support that decreases slightly for the non-meritocratic ones. Still, we have to consider that the preferences for non-meritocratic aspects in this case refer to the distributive principle of need rather than personal background (as in the perceptions of non-meritocracy), which certainly is one of the limitations of this set of items for the operationalization of the proposed conceptual framework with this dataset.

@@

First, a four-dimensional model for meritocratic and non-meritocratic perception and preferences is estimated, considering all indicators of the non-meritocratic perception dimension as a single latent variable. Then, we estimate a model that follows the proposed four latent variable measurement approach, however, the non-meritocratic perception dimension is assumed as a second-order factor for the first order latent variables parental attributes, racial and gender background and networks.

@@
